\documentclass[12pt,a4paper]{article}
\input{preamble}


% \addbibresource{references.bib} % Note the .bib extension is required here

% --- Document Metadata ---
\title{
    % \vspace{5em}
    \large \textbf{CPSC 704 - Graduate Seminar in Computer Science} \\

    \huge \textbf{Paper Critique} \\

    \vspace{0.5em}

    \large \textit{Quantifying Attention Flow in Transformers}
}

\author{
    \large Submitted by \\[0.2cm]
    \renewcommand{\arraystretch}{1.5}
    \begin{tabular}{|c|c|}
        \hline
        \textbf{Student ID} & \textbf{Name} \\
        \hline
        230171256 & Simon Kraft \\
        \hline
    \end{tabular}
}

\date{March 18, 2026}

\begin{document}

\maketitle
\newpage

\section{Summary}

The work \citetitle{abnar-zuidema-2020-quantifying} by~\citeauthor{abnar-zuidema-2020-quantifying}~\cite{abnar-zuidema-2020-quantifying} proposes two approaches to quantify and visualize the raw attention weights of Transformer models as explanations for the model's decisions. \\

Usually, the self-attention mechanism in Transformer models is used so that the information of multiple tokens can get ''mixed'', i.e. the representation of each token captures information, also called \textit{context}, from all other processed tokens. While this mechanism is really powerful to model long-range dependencies between tokens, it complicates interpretability. As information propagates through successive layers, the attention weights, i.e. the amount how much one token's representation influences another's token representation, becomes harder to understand, as the hidden representation of the token already contains a mixture of the entire token sequence. \\

To overcome this, the authors model the information flow between tokens over multiple layers as a Directed Acyclic Graph (DAG), in which the nodes are the input tokens or the hidden embeddings, and the edges model the attention weights from nodes in each layer to those representations in the previous layer. Based on this DAG, they introduce two distinct methods:

\begin{itemize}
    \item \textbf{Attention Rollout:} This method assumes that input token identities are linearly combined across layers. It recursively multiplies the attention matrices from the input layer to the target layer. This effectively sums the products of edge weights along all possible paths between two nodes.
    \item \textbf{Attention Flow:} This method treats the attention graph as a flow network where attention weights act as edge capacities. By applying a maximum flow algorithm, it computes the total information flow from hidden embeddings to input tokens. This provides a more strict distinction between tokens who constantly had a high attention score and tokens who maybe only had a high attention score in one of the layers.
\end{itemize}


\section{Shortcomings}

One shortcoming of the work is the relatively narrow scope of the experimental evaluation. While the methods show high correlation with ground-truth importance scores on the syntactic task of verb number prediction, they perform significantly worse on semantic tasks like sentiment analysis (SST-2). In the latter case, the proposed methods showed low correlation with input gradients, this suggests that they may not fully capture the dependencies within large-scale, fine-tuned models.

Furthermore, the authors did not apply their method on datasets beyond the NLP domain. 

\section{Flaws in Theory}

The theoretical framework of Attention Rollout and Attention Flow relies on the assumption that the non-linear activations functions and the linear transformations that are applied in the position-wise feed-forward network in each transformer layer can be ignored. Furthermore, the layer normalizations in each transformer layer are ignored. \\

In their modeling approach, the authors treat the attention weights as the sole proxy for information transfer within a transformer network. This may not be the case, especially not when transformer networks become more deep and have multiple successive layers.


\section{Unanswered Questions}

The following questions regarding the scalability and generalizability remain unanswered by the paper:
\begin{itemize}
    \item How do these methods scale with much deeper architectures, with e.g. 24 or 48 layers, where the "mixing" of tokens is significantly more extreme?
    \item How can these methods be adapted for Transformer decoders with causal masking, which naturally biases attention toward early tokens in a sequence?~\cite{abnar-zuidema-2020-quantifying}
    \item How can these methods be applied to other domains like Computer Vision?
    \item Can token attention scores be used during training (e.g., as a regularizer) to actively encourage more interpretable information flow, rather than just being used as a post-training diagnostic tool?
\end{itemize}

\section{Future Research}

The authors propose different avenues for further research. For instance, one approach could be to combine the attention graph with gradient information to build a more accurate flow of attention instead of solely relying on the raw attention weights for the explanations. Furthermore, the authors invite their audience to think about ways to adapt their method to transformer decoder models. These decoder models use causal masking which usually leads to a more focused attention towards the earlier tokens in a sequence. Lastly, one could examine ways to integrate the effects of the position-wise feed-forward networks to the explanation strategy.



\end{document}